
---
title: "Reproductible Research Course - Project 1"
date: "Wednesday, July 15, 2015"
output:
  html_document:
    keep_md: true
---

This project answers questions from Peer Assessment asignment #1, of the reproductible Research course on Coursera.


<hr>

## Loading and preprocessing the data

<hr>
#### Load the data (i.e. read.csv()) and process/transform the data (if necessary) into a format suitable for your analysis

The code below will load all of the libraries required by this project, handle some global formatting issues and finally load the dataset.

```{r, message=FALSE}

## Setting up the library and folder used in the assignment
library(dplyr)
library(ggplot2)
library(xtable)
setwd("C:/Users/jfrue_000.MAISON/R/Course5/RepData_PeerAssessment1")
# setwd("~/R")

## numbers >= 10^5 will be denoted in scientific notation and rounded to 2 digits
options(scipen = 1, digits = 2)

## Reads the file in data frames that can be used by dplyr
activities <- tbl_df(read.csv("activity.csv"))
```

<hr>

## What is mean total number of steps taken per day?

<hr>
For this part of the assignment, you can ignore the missing values in the dataset.

#### Calculate the total number of steps taken per day

```{r, results='asis'}
# Creates the summary by date, and filters out the "NA"s
SPD <- summarise(group_by(filter(activities,steps != "NA"),date),Total_Steps=sum(steps))

#Use the xtable library to create an html table with all the results
print(xtable(SPD, caption="<b>Total</b> number of steps taken per day"), type="html", caption.placement="top")
```
<br><br>

#### Make a histogram of the total number of steps taken each day

```{r}
ggplot(SPD, aes(x=Total_Steps)) +
  geom_histogram(fill = "blue", colour="pink", binwidth=2500) +
  labs(title="Histogram of number of days by range of steps taken\n",
       x = "\nTotal number of steps taken in a given day", y = "Frequency of days\nfor each steps range") +
  theme_bw()

```

#### Calculate and report the mean and median of the total number of steps taken per day

```{r}
# creates the mean and median variables to be
# inserted in text using the `R [...]` function
mean_SPD <- mean(SPD$Total_Steps)
Median_SPD <- median(SPD$Total_Steps)


```

The <b>mean</b> of the total number of steps per day is <b>`r mean_SPD`</b>

The <b>median</b> of the total number of steps per day is <b>`r Median_SPD`</b>

<hr>

## What is the average daily activity pattern?

<hr>

- Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
- Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
